Using Genetic Algorithms to Optimize the Analogue Method for Precipitation Prediction in the Swiss Alps

### 2016-12-22 Optimization case study - JoH/elsarticle-template.tex
JoH paper under correction.

@@

\address[unil]{University of Lausanne, Institute of Earth Sciences, Lausanne, Switzerland}
\address[unibe]{University of Bern, Oeschger Centre for Climate Change Research, Institute of Geography, Bern, Switzerland}
\address[terranum]{Terranum SARL, Bussigny, Switzerland}
\address[lthe]{Universit\'{e} de Grenoble-Alpes, LTHE, Grenoble, France}

\begin{abstract}
Analogue methods aim at predicting precipitation based on synoptic predictors provided by general circulation models or numerical weather prediction models. The method samples a selection of days in the archives that are similar to the target day to predict, and consider the set of their corresponding observed precipitation (the predictand) as the conditional distribution for the target day. The relationship between predictors and predictands relies on some parameters that characterize how and where the similitude between two atmos+pheric situations is defined.

This relationship is usually established by a semi-automatic sequential procedure that has strong limitations: (i) it cannot automatically choose the pressure levels and the temporal windows for a given meteorological variable, (ii) it cannot handle dependencies between the parameters within a level of analogy, and even less between them, and (iii) it could not easily handle new degrees of freedom. A new global optimization approach relying on genetic algorithms can optimize all parameters jointly and automatically, which is a breakthrough in the way the analogue method was calibrated until now. It allows taking into account parameters inter-dependencies, and selecting objectively some parameters that were manually selected beforehand (such as the pressure levels and the temporal windows of the predictor variables).

In this work, the global optimization is applied to the Rh\^{o}ne catchment, in the Swiss Alps. The performance scores are significantly increased compared to a reference method, and this even to a greater extent for days with high precipitation totals. The resulting parameters were found to be relevant and coherent between the different subregions of the catchment. Moreover, they are obtained automatically and objectively, which reduces efforts invested in exploration attempts when adapting the method to a new region or for a new predictand. In addition, the approach allows for new degrees of freedom, such as a weighting between the pressure levels, and non overlapping spatial windows.
\end{abstract}

\begin{keyword}
Precipitation prediction\sep
Precipitation downscaling\sep
Analogue method\sep
Optimization\sep
Genetic algorithms\sep
Alpine climate
\end{keyword}

@@

The analogue method (AM) is a downscaling technique based on the idea expressed by \citet{Lorenz1956, Lorenz1969}, that similar situations in terms of atmospheric circulation are likely to lead to similar local weather \citep{Duband1970}. It uses predictor variables describing the synoptic atmospheric circulation in order to predict local-scale predictands of interest. It is often used to predict daily precipitation, either in an operational forecasting context \citep[e.g.,][]{Guilbaud1997, Bontron2005, Hamill2006, Bliefernicht2010, Marty2012, Horton2012, Hamill2015, BenDaoud2016, Horton2016a}, or a climate context \citep[e.g.,][]{Radanovics2013, Chardon2014, Dayon2015, Raynaud2016b}. Other predictands are also considered, such as precipitation radar images \citep{Panziera2011,Foresti2015a}, temperature \citep{Radinovic1975, Woodcock1980, Kruizinga1983, DelleMonache2013, Caillouet2016, Raynaud2016b}, wind \citep{Gordon1987, DelleMonache2013, DelleMonache2011, Vanvyve2015, Alessandrini2015, Junk2015, Junk2015c}, solar power \citep{Alessandrini2015a, Bessa2015}, snow avalanches \citep{Obled1980, Bolognesi1993}, and radiation \citep{Bois1981, Raynaud2016b}.

In operational forecasting, it has been used mainly by practitioners, notably hydropower companies or flood forecasting services, while in climate studies, it is used to downscale the outputs of general circulation model (GCM) or regional climate models (RCMs) simulation runs \citep{Dayon2015} or to reconstruct past weather conditions \citep{Caillouet2016}. In future climate studies, RCMs are often used to dynamically downscale precipitation to a local scale. However, even though the relevance of RCMs'outputs increases, precipitation processes are parametrized and a bias correction of the outputs is still often required. Moreover, their application is computer intensive and cannot cover the full range of GCMs products. Therefore the idea is to rely on past observations to bypass the small scale simulations and to go from the large scale situation to the end variables like precipitation by searching analogues in the archives.

@@

The method can be designed with multiple successive subsampling steps, each of them relying on different meteorological variables. A certain number of parameters define the relationship between predictors and predictands, such as the choice of the predictor variable, its pressure level and temporal window to consider, the spatial domain to use for the comparison, as well as the analogy criterion itself, and finally the number of analogue situations to keep at each subsampling level. These parameters are usually calibrated by means of a semi-automatic sequential procedure \citep[see][for the details]{Bontron2004, Radanovics2013}, i.e. by optimizing each single parameter one at a time, in an arbitrarily chosen order, with no or little reassessment. This sequential approach has therefore strong limitations: (i) it cannot automatically choose the optimal pressure levels and the temporal windows for a given meteorological variable, (ii) it cannot handle dependencies between the parameters within a level of analogy, and even less between them, and (iii) it could not easily handle new degrees of freedom, such as a weighting between the pressure levels. Thus, even if the processing involved is relatively fast, the sequential approach requires laborious assessments of predictors combinations (variables, pressure levels, temporal windows), and presents a high risk of ending in a local optimum due to subjective initial choices and lack of consideration of parameter inter-dependencies. Other calibration methods exist for specific applications, such as the radar images \citep{Panziera2011, Foresti2015a}.

With the perspective to overcome these limitations, a global optimization by genetic algorithms (GAs) was introduced. An intensive assessment work resulted in recommendations of parametrization of GAs in order to optimize AMs successfully \citet{Horton2016}. The present paper is based on these recommendations and apply them for precipitation prediction for the upper Rh\^{o}ne catchment in the Swiss Alps. It aims at proving the relevance of a fully-automatic, objective, and global, optimization technique for AMs. The applications are indeed numerous, as the AM has to be adapted to every new location it is applied, or to any new predictand it should predict.

A short overview of AMs is presented in section \ref{sec:data_methods}, as well as a summary of genetic algorithms in section \ref{sec:gas}. Section \ref{sec:case_study} describes the case study area. The results are first detailed for the optimization of the analogy on the atmospheric circulation only (section \ref{sec:optim_circul}), before being extended to a method adding a second level of analogy on moisture variables (section \ref{sec:optim_moisture}). General discussions (section \ref{sec:discussion}) and conclusions (section \ref{sec:conclusions}) follow.

@@

The study area is the alpine upper Rh\^{o}ne catchment in Switzerland (Fig.\ \ref{fig:map}). The altitude ranges from 372 to 4634~m.a.s.l.\ and the area is 5524~km$^{2}$. This region is the target of the MINERVE (Mod\'{e}lisation des Intemp\'{e}ries de Nature Extr\^{e}me sur les Rivi\`{e}res Valaisannes et de leurs Effets) project that aims at allowing real-time flood management on the upper Rh\^{o}ne catchment \citep{GarciaHernandez2009b}. Even though the region is rather small, the meteorological influences related to extreme weather conditions varies substantially within it \citep[see][]{Horton2012}. Indeed, a high spatial variability exists, which is due to the complex orography of the region. Based on different climatological analyses, the gauging stations in the catchment were clustered in 10~subregions (Fig.\ \ref{fig:map}) :

@@

The 48 years precipitation dataset (see section \ref{sec:intro}.\ref{sec:data}) was divided into a calibration period (CP) and a validation period (VP). Using data independent from the CP to validate the results is very important in order to assess the robustness of the improvements and to avoid over-parametrization of the method. Parameters determined on the CP are then applied to the VP in order to obtain a performance score for the independent period.

@@

The AM relies on two types of data: predictors, that are atmospheric variables describing the state of the atmosphere at a synoptic scale, and the predictand, which is the local weather time series one wants to predict.

@@

>>>>

%TODO: read again
%TODO: tenses
%TODO: Elsevier english editing
%TODO: more hydrology

@@

\address[unil]{University of Lausanne, Institute of Earth Sciences, Lausanne, Switzerland}
\address[unibe]{University of Bern, Oeschger Centre for Climate Change Research, Institute of Geography, Bern, Switzerland}
\address[lthe]{Universit\'{e} de Grenoble-Alpes, LTHE, Grenoble, France}

\begin{abstract}
Analogue methods provide a statistical precipitation prediction based on synoptic predictors supplied by general circulation models or numerical weather prediction models. The method samples a selection of days in the archives that are similar to the target day to predict, and consider the set of their corresponding observed precipitation (the predictand) as the conditional distribution for the target day. The relationship between predictors and predictands relies on some parameters that characterize how and where the similitude between two atmospheric situations is defined.

This relationship is usually established by a semi-automatic sequential procedure that has strong limitations: (i) it cannot automatically choose the pressure levels and the temporal windows for a given meteorological variable, (ii) it cannot handle dependencies between the parameters within a level of analogy, and even less between them, and (iii) it cannot easily handle new degrees of freedom. A new global optimization approach relying on genetic algorithms can optimize all parameters jointly and automatically, which is a breakthrough in the way the analogue method was calibrated until now. It allows taking into account parameters inter-dependencies, and selecting objectively some parameters that were manually selected beforehand (such as the pressure levels and the temporal windows of the predictor variables).

In this work, the global optimization was applied to the Rh\^{o}ne catchment in the Swiss Alps. The performance scores increased compared to reference methods, and this even to a greater extent for days with high precipitation totals. The resulting parameters were found to be relevant and coherent between the different subregions of the catchment. Moreover, they were obtained automatically and objectively, which reduces efforts invested in exploration attempts when adapting the method to a new region or for a new predictand. In addition, the approach allowed for new degrees of freedom, such as a weighting between the pressure levels, and non overlapping spatial windows.
\end{abstract}

\begin{keyword}
precipitation prediction\sep
analogue method\sep
optimization\sep
genetic algorithms\sep
Alpine climate
\end{keyword}

@@

The analogue method (AM) is a downscaling technique based on the idea expressed by \citet{Lorenz1956, Lorenz1969} that similar situations in terms of atmospheric circulation are likely to lead to similar local weather \citep{Duband1970}. It uses predictor variables describing the synoptic atmospheric circulation in order to predict local-scale predictands of interest. It is often used to predict daily precipitation, either in an operational forecasting context \citep[e.g.,][]{Guilbaud1997, Bontron2005, Hamill2006, Bliefernicht2010, Marty2012, Horton2012, Horton2016a, Hamill2015, BenDaoud2016} or a climate context \citep[e.g.,][]{Radanovics2013, Chardon2014, Dayon2015, Raynaud2016b}. Other predictands are also considered, such as precipitation radar images \citep{Panziera2011,Foresti2015a}, temperature \citep{Radinovic1975, Woodcock1980, Kruizinga1983, DelleMonache2013, Caillouet2016, Raynaud2016b}, wind \citep{Gordon1987, DelleMonache2013, DelleMonache2011, Vanvyve2015, Alessandrini2015, Junk2015, Junk2015c}, solar power \citep{Alessandrini2015a, Bessa2015}, snow avalanches \citep{Obled1980, Bolognesi1993}, and radiation \citep{Bois1981, Raynaud2016b}.

In operational forecasting it has been used mainly by practitioners, notably hydropower companies or flood forecasting services, while in climate studies it is used to downscale the outputs of general circulation model (GCM) or regional climate models (RCMs) simulation runs \citep{Dayon2015} or to reconstruct past weather conditions \citep{Caillouet2016}. In future climate studies, RCMs are often used to dynamically downscale precipitation to a local scale. However, even though the relevance of RCMs'outputs increases, a bias correction of the outputs is still often required, particularly in complex terrain. Moreover, their application is computer intensive, which makes it difficult to cover all combinations of climate scenarios and GCMs. Therefore, the idea is to rely on past observations to bypass the small scale simulations and to go from the large scale situation to the end variables like precipitation by searching analogues in the archives.

@@

The method can be designed with multiple successive subsampling steps, each of them relying on different meteorological variables. A certain number of parameters define the relationship between predictors and predictands, such as the choice of the predictor variable, its pressure level and temporal window to consider, the spatial domain to use for the comparison, as well as the analogy criterion itself, and finally the number of analogue situations to keep at each subsampling level. These parameters are usually calibrated by means of a semi-automatic sequential procedure \citep{Bontron2004, Radanovics2013}, i.e. by optimizing each single parameter one at a time, in an arbitrarily chosen order, with no or little reassessment. This sequential approach has therefore strong limitations: (i) it cannot automatically choose the optimal pressure levels and the temporal windows for a given meteorological variable, (ii) it cannot handle dependencies between the parameters within a level of analogy, and even less between them, and (iii) it cannot easily handle new degrees of freedom, such as a weighting between the pressure levels. Thus, even if the processing involved is relatively fast, the sequential approach requires laborious assessments of predictors combinations (variables, pressure levels, temporal windows), and presents a high risk of ending in a local optimum due to subjective initial choices and lack of consideration of parameter inter-dependencies. Other calibration methods exist for specific applications, such as radar images \citep{Panziera2011, Foresti2015a}.

With the perspective to overcome these limitations, a global optimization by genetic algorithms (GAs) was introduced. An intensive assessment work resulted in recommendations of parametrization of GAs in order to optimize AMs successfully \citep{Horton2016}. The present paper is based on these recommendations and apply them for precipitation prediction for the upper Rh\^{o}ne catchment in the Swiss Alps, using AMs of varying complexity. It aims at illustrating the relevance of a fully-automatic, objective, and global, optimization technique for AMs. The applications are indeed numerous, as AMs have to be adapted to every new location they are applied, or to any new predictand they should predict.

Data, AMs, and the optimization techniques (sequential and GAs) are presented in section \ref{sec:data_methods}. The results are first given for the optimization of the analogy of the atmospheric circulation only (section \ref{sec:optim_circul}), before being extended to a method adding a second level of analogy on moisture variables (section \ref{sec:optim_moisture}). General discussions (section \ref{sec:discussion}) and conclusions (section \ref{sec:conclusions}) follow.

@@

The study area is the alpine upper Rh\^{o}ne catchment in Switzerland (Fig.\ \ref{fig:map}). The altitude ranges from 372 to 4634~m.a.s.l.\ and the area is 5524~km$^{2}$. This region is the target of the MINERVE (Mod\'{e}lisation des Intemp\'{e}ries de Nature Extr\^{e}me sur les Rivi\`{e}res Valaisannes et de leurs Effets) project that aims at real-time flood management on the upper Rh\^{o}ne catchment \citep{GarciaHernandez2009b}. Even though the region is rather small, the meteorological influences related to extreme weather conditions vary substantially within it \citep[see][]{Horton2012}. Indeed, a high spatial variability exists, which is due to the complex orography of the region, and the mix of various meteorological influences. Based on different climatological analyses, the gauging stations in the catchment were clustered in 10~subregions (Fig.\ \ref{fig:map}) :

@@

@@

The AM relies on two types of data: predictors, that are atmospheric variables describing the state of the atmosphere at a synoptic scale, and the predictand, which is the local weather variable one wants to predict.

@@

The 48 years precipitation dataset (see section \ref{sec:data}) was divided into a calibration period (CP) and a validation period (VP). Using data independent from the CP to validate the results is very important in order to assess the robustness of the proposed improvements and to avoid over-parametrization of the method.
